Formalizing Abstract Rewriting

### 2025-02-28 Relations/Paper/FinalTODO.tex
updating ARS branch

@@

>>>>

\documentclass{scrartcl}

\input{packages.tex}
\input{macros.tex}

\begin{document}

\section{Feb 28 meeting}

\begin{enumerate}
  \item Add the implication RisWCR, x in WN and SM implies x in SN to Section 4.3.
  \item Clean up [TeReSe] references.
  \item Clean up the Well-foundedness
  \item Put up all classical principles in one place
  \item Organize the repo, hide private work.
  \item Clean up the Implications file, rename section etc.
  \item Getting rid of confluent/weakly confluent definition, standardize
\end{enumerate}

For the long term.
\begin{itemize}
  \item Knaster--Tarski, look over that
  \item Kleene tree to disprove isWFseq implies isWFacc
\end{itemize}

@@

\end{document}

### 2025-02-28 Relations/notes.tex
updating ARS branch

\usepackage{xcolor}

@@

\section*{Notes}

@@

Also discuss open problems, unprovable implications, etc.
Maybe mention the Kleene tree idea, or at least include the reference to the Bezem-Coquand paper.

\section{Motivation}

The standard presentation in Terese makes liberal use of classical logic.
Should stress effectivity of type-theoretic formalization.
\section{Contributions}

\begin{itemize}
  \item Formalization of ``elementary ARS theory'' as presented in
  Chapter 1 of [TeReSe].
  \item An ontology of termination and confluence properties, and a detailed analysis of logical relations between them.
  \item Defining several new ARS concepts that serve to streamline and generalize a number of relationships.
  \item With the above, obtain marginal improvements to classical confluence and termination criteria
  \item Collecting implications between constructive notions of well-foundedness
\end{itemize}

\section{Formalization principles}

@@

One big item in this development was studying various incaranations of the concept of well-foundedness in constructive context.
(Strong normalization is well-foundedness of the converse relation.)
The ``standard'' constructive definition does not even allow one to show that SN implies WN, see Example \ref{ex:undec}.  We therefore looked at a number of variations of this notion, their classical counterparts all being logically equivalent.  This gave rise to a rich set of concepts, see Figure \ref{f:wf}.

>>>>

\usepackage[dvipsnames]{xcolor}

@@

\section{Final notes about well-foundedness}

\begin{enumerate}
  \item The implication acc $\to$ seq requires only decidability of the relation itself.
  \item Finitely branching and decidable implies that they're all equivalent, almost.

\end{enumerate}

@@

Also discuss open problems, unprovable implications, etc.
Maybe mention the Kleene tree idea, or at least include the reference to the Bezem--Coquand paper.

\section{Introduction}
We present a formalization of the basic Abstract Rewriting Systems (ARS) theory in the Agda proof assistant.
This is part of a larger effort to develop a library of formalized programming language theory (PL)
at Appalachian State University.
Since Term Rewriting Systems (TRS) play a foundational role in programming languages,
while ARS encompass the basic facts about rewriting relations in general,
establishing these facts is part of the necessary infrastructure required to
pursue the broader project.  The present contribution can therefore be seen as
``bootstrapping'' Appstate's new formalized PL repository.

This broader goal also drives the main design choices in our approach.
Since rewriting theory is concerned with a fine-grained analysis of computation,
the most natural vehicle for formalizing it is type theory based on the
Curry--Howard isomorphism, which Agda implements.  Proofs in this language
are automatically effective, and implicitly carry the code implementing the
transformations from the hypotheses to the conclusion.  For example,
a proof that a given TRS is terminating automatically renders a function that
computes a normal form of any term in its domain.

Staying within this paradigm requires the proofs to be constructive.
Moreover, it requires us to avoid any axioms or postulates, such as function
extensionality, uniqueness of identity proofs, univalence, etc.
While most of ARS results are indeed constructive, standard presentations,
 including \hired{[Terese]}, make liberal use of classical logic.
Our paper can thus be read as a thoroughly constructive
development of elementary ARS theory.

Before we proceed further, let us consider two formalization scenarios where
our contribution can be useful.

\subsubsection*{Example 1. Formalization of typed lambda calculus.}

Suppose one is formalizing the standard metatheory of
a typed lambda calculus.  To show confluence, one possible approach is to
first establish the Church--Rosser theorem for the untyped lambda calculus,
then invoke the Subject Reduction Theorem.
However, suppose one has already independently verified Strong Normalization.
Invoking Newman's lemma, one can then conclude confluence from the weak
diamond property, which is generally a much simpler fact to prove.

\subsubsection*{Example 2. Quotients of types.}
Suppose one wishes to construct the initial model of an algebraic theory.
The classical definition involves the quotient of a free algebra by
the equivalence relation induced by the theory.
The quotient is problematic operation in type theory.
While there are various solutions, such as Higher Inductive Types,
or Quotient Inductive--Inductive Types, these require one to extend the
basic type theory with new constructs.

However, if the algebraic theory in question admits a convergent presentation,
by a confluent and terminating relation $R : A \to A \to Set$, one can define the type of
$R$-normal forms as
\[  (\Sigma n : A) (\Pi x : A) \lnot P n x \]
In many applications, this type can substitute the true quotient satisfactorily.
(In univalent or cubical type theories, one can moreover show this type to be
a set, provided the domain $A$ is.)

As these examples illustrate, a fully effective implementation of basic ARS results
can be very helpful in many settings arising in programming language theory.

\section{Contributions}
Our contributions
\begin{itemize}
  \item Formalization of ``elementary ARS theory'' as presented in
  Chapter 1 of [TeReSe].
  \item An ontology of termination and confluence properties, and a detailed analysis of logical relations between them.
  \item Defining several new ARS concepts that serve to streamline and generalize a number of relationships.
  \item With the above, obtain marginal improvements to classical confluence and termination criteria
  \item Collecting implications between constructive notions of well-foundedness
\end{itemize}

\subsection{Formalization principles}
\hired{One of our design goals is to stay within pure type theory as much as possible,
therefore we adopt the following principles.}

@@

\subsection{Misc.}
One big item in this development was studying various incaranations of the concept of well-foundedness in constructive context.
(Strong normalization is well-foundedness of the converse relation.)
The ``standard'' constructive definition does not even allow one to show that SN implies WN, see Example \ref{ex:undec}.  We therefore looked at a number of variations of this notion, their classical counterparts all being logically equivalent.  This gave rise to a rich set of concepts, see Figure \ref{f:wf}.
